Testing spacetime model with Kronecker sums on WAIS data

### 2022-07-20 notebooks/deprecated/kronecker-tests.Rmd
Cleans up old notebook files

set.seed(777)
gdf_traces = gdf_traces %>% sample_frac(0.01)

# # Subset to smaller region
# tmp = st_bbox(gdf_traces)
# East = c(tmp[1] + (tmp[3]-tmp[1])/7, tmp[1] + (tmp[3]-tmp[1])/2)
# North = c(tmp[2] + (tmp[4]-tmp[2])/2, tmp[4] - (tmp[4]-tmp[2])/8)
# tmp.df = tibble(East=East, North=North)
# poly = tmp.df %>%
#   st_as_sf(coords = c("East", "North"),
#            crs = 3031) %>%
#   st_bbox() %>%
#   st_as_sfc()
# gdf.idx = sapply(st_intersects(gdf_traces, poly),function(x){length(x)==0})
# gdf_traces = gdf_traces[!gdf.idx,]

@@

```{r}
# More coarse mesh for development purposes
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(60000, 120000), cutoff = 12000)
# mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
#                     max.edge = c(30000, 90000), cutoff = 1000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")

@@

@@

# mod.gamma = inla(form,
#                data = inla.stack.data(dat.stack),
#                family = 'Gamma',
#                control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
#                control.inla = list(int.strategy='eb', strategy = 'gaussian'), # Options improve computation time at cost of accuracy/precision
#                control.compute = list(waic=TRUE, config=TRUE))
mod.gamma = readRDS(here('data/interim_results/mod-gamma.rds'))

summary(mod.gamma)
```

@@

It's possible that I am removing temporal signal from my estimates when I model the spacetime auto-correlations.
Here I run the model without this random effect for comparison (so only using a spatially varying temporal coefficient).

```{r}
# Model formula
f.noST = y ~ -1 + Intercept + 
  # f(st.idx, model = spde, group=st.idx.group, 
  #   control.group=list(model='ar1')) + 
  f(yr.idx, model = spde)

mod.noST = inla(f.noST,
                data = inla.stack.data(dat.stack),
                family = 'Gamma',
                control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
                control.inla = list(int.strategy='eb', strategy = 'gaussian'), # Options improve computation time at cost of accuracy/precision
                control.compute = list(waic=TRUE, config=TRUE))

summary(mod.noST)
```

@@

```{r}
# Posterior draws for year coefficients
draws=100
n.locs = spde$n.spde
samples.noST = inla.posterior.sample(draws, result = mod.noST)
tmp = inla.posterior.sample.eval(c("yr.idx"), samples.noST)
ppc.noST = tibble(Sample=rep(1:draws, each=n.locs), coeff=as.vector(tmp))

ggplot(ppc.noST %>% group_by(Sample)) + 
  geom_line(aes(x=coeff, group=Sample), stat="density", alpha=0.1) + 
  geom_density(data=dat.mu, aes(x=log.est), color='red') + 
  xlim(c(-0.03, .03))
```

@@

```{r}
tmp = tibble(Source="model", 
             East=mesh$loc[,1], North=mesh$loc[,2], 
             med=mod.noST$summary.random$yr.idx$`0.5quant`, 
             LB=mod.noST$summary.random$yr.idx$`0.025quant`, 
             UB=mod.noST$summary.random$yr.idx$`0.975quant`) %>% 
  # mutate(weak=(LB<0 && UB>0)) %>% 
  st_as_sf(coords=c("East", "North"), crs=3031)

trends.noST = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
  rename(med=log.est) %>% mutate(LB=NA, UB=NA, Source="data") %>% 
  st_as_sf(coords=c("East", "North"), crs=3031) %>%
  bind_rows(tmp)

ggplot(trends.noST %>% filter(Source=="data")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.03, 0.03))
ggplot(trends.noST %>% filter(Source=="model")) + 
  geom_sf(aes(color=LB)) + scale_color_gradient2(limits=c(-0.03, 0.03))
ggplot(trends.noST %>% filter(Source=="model")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.03, 0.03))
ggplot(trends.noST %>% filter(Source=="model")) + 
  geom_sf(aes(color=UB)) + scale_color_gradient2(limits=c(-0.03, 0.03))
```

>>>>

# # Subset to smaller region
# tmp = st_bbox(gdf_traces)
# # East = c(tmp[1] + (tmp[3]-tmp[1])/7, tmp[1] + (tmp[3]-tmp[1])/2)
# # North = c(tmp[2] + (tmp[4]-tmp[2])/2, tmp[4] - (tmp[4]-tmp[2])/8)
# East = c(tmp[1] + (tmp[3]-tmp[1])/7, tmp[3])
# North = c(tmp[2], tmp[2] + (tmp[4]-tmp[2])/5)
# tmp.df = tibble(East=East, North=North)
# poly = tmp.df %>%
#   st_as_sf(coords = c("East", "North"),
#            crs = 3031) %>%
#   st_bbox() %>%
#   st_as_sfc()
# ggplot() + geom_sf(data=poly) + geom_sf(data=gdf_traces)
# gdf.idx = sapply(st_intersects(gdf_traces, poly),function(x){length(x)==0})
# gdf_traces = gdf_traces[!gdf.idx,]

set.seed(777)
# gdf_traces = gdf_traces %>% sample_frac(0.05)
skip.int = 5
gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
gdf_traces = gdf_traces[gdf.idx,]

@@

```{r}
# More coarse mesh for development purposes
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
                    max.edge = c(60000, 120000), cutoff = 12000)
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
                    max.edge = c(30000, 90000), cutoff = 2000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")

@@

Indexing the spacetime function by year directly, rather than having two seperate terms for it (I'm not sure this will work because we don't have unique year values for each spacetime point).
```{r}
A.st <- inla.spde.make.A(mesh, 
                         loc=cbind(dat.train$East, dat.train$North), 
                         group = dat.train$Year.idx, 
                         weights = dat.train$Year.mod)

dat.stack <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, A.st),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        # slope=dat.train$slope, aspect=dat.train$aspect, 
                        Year=dat.train$Year.mod),
                 st.idx),
  tag = 'dat')

form = y ~ -1 + Intercept + #modeling the intercept
  f(st.idx, model = spde, group=st.idx.group, 
    control.group=list(model='ar1'))
```

@@

mod.gamma = inla(form,
               data = inla.stack.data(dat.stack),
               family = 'Gamma',
               control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
               # control.inla = list(strategy = 'simplified.laplace', int.strategy='ccd'), # Options improve computation time at cost of accuracy/precision
               control.inla = list(strategy = 'gaussian', int.strategy='eb'), 
               control.compute = list(waic=TRUE, config=TRUE))

# mod.gamma = readRDS(here('data/interim_results/mod-gamma.rds'))
summary(mod.gamma)
```

@@

It's possible that I am removing temporal signal from my estimates when I model the spacetime auto-correlations.
Here I run the model without this random effect for comparison (so only using a spatially varying temporal coefficient and a separate temporal random effect).

```{r}
# Make data stack
stack.sep <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, 1, A.yr),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect, 
                        Year=dat.train$Year),
                 list(time = dat.train$Year),
                 yr.idx),
  tag = 'dat')

# Priors for temporal autocorrelation
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))

# Model formula
f.sep = y ~ -1 + Intercept + 
  # f(st.idx, model = spde, group=st.idx.group, 
  #   control.group=list(model='ar1')) + 
  f(yr.idx, model = spde) + 
  f(time, model = 'ar1', hyper = time.spec) #Temporal random effect modeled as AR1

mod.sep = inla(f.sep,
                data = inla.stack.data(stack.sep),
                family = 'Gamma',
                control.predictor = list(compute = TRUE, A = inla.stack.A(stack.sep), link=1),
                control.inla = list(int.strategy='eb'), # Option improve computation time at cost of accuracy/precision
                control.compute = list(waic=TRUE, config=TRUE))

summary(mod.sep)
```

@@

Extract marginal distributions for subset of coefficients (for plotting purposes)
```{r}
n=50
idx.set = as.integer(seq(1,spde$n.spde, length.out=n))
marg.set = mod.sep$marginals.random$yr.idx[idx.set]

# Create tbl of marginals
marg.tbl = tibble()
for (i in 1:length(idx.set)) {
  set.i = marg.set[[i]]
  tbl.i = tibble(Index = idx.set[i], X=set.i[,1], Y=set.i[,2])
  marg.tbl = marg.tbl %>% bind_rows(tbl.i)
}

ggplot(marg.tbl) + geom_line(aes(x=X, y=Y, group=Index), color='black', alpha=0.1)
```

Calculate probability that trend is less than zero.
```{r}
P.neg = vector(mode = "numeric", length = length(mod.sep$marginals.random$yr.idx))
for (i in 1:length(P.neg)) {
  marg.i = inla.tmarginal(function(x) x, 
                          mod.sep$marginals.random$yr.idx[[i]], 
                          n=500)
  P.i = diff(marg.i[,1]) * marg.i[1:(nrow(marg.i)-1),2]
  P.neg[i] = sum(P.i[which(marg.i[1:(nrow(marg.i)-1),1]<0)])
}
```

Map of estimated trends with probability of less than zero
```{r}
tmp = tibble(Source="model", 
             East=mesh$loc[,1], North=mesh$loc[,2], 
             med=mod.sep$summary.random$yr.idx$`0.5quant`, 
             P.neg=P.neg, 
             LB=mod.sep$summary.random$yr.idx$`0.025quant`, 
             UB=mod.sep$summary.random$yr.idx$`0.975quant`) %>% 
  # mutate(weak=(LB<0 && UB>0)) %>% 
  st_as_sf(coords=c("East", "North"), crs=3031)

trends.sep = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
  rename(med=log.est) %>% mutate(LB=NA, UB=NA, P.neg=NA, Source="data") %>% 
  st_as_sf(coords=c("East", "North"), crs=3031) %>%
  bind_rows(tmp)

ggplot(trends.sep %>% filter(Source=="data")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.05, 0.05))
# ggplot(trends.sep %>% filter(Source=="model")) + 
#   geom_sf(aes(color=LB)) + scale_color_gradient2(limits=c(-0.05, 0.05))
ggplot(trends.sep %>% filter(Source=="model")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.05, 0.05))
# ggplot(trends.sep %>% filter(Source=="model")) + 
#   geom_sf(aes(color=UB)) + scale_color_gradient2(limits=c(-0.05, 0.05))
ggplot(trends.sep %>% filter(Source=='model')) + 
  geom_sf(aes(color=P.neg)) + scale_color_viridis_c()
```

```{r}
# Posterior draws for year coefficients
draws=10
n.locs = spde$n.spde
samples.sep = inla.posterior.sample(draws, result = mod.sep)
tmp = inla.posterior.sample.eval(c("yr.idx"), samples.sep)
ppc.sep = tibble(Sample=rep(1:draws, each=n.locs), coeff=as.vector(tmp))

ggplot(ppc.sep %>% group_by(Sample)) + 
  geom_line(aes(x=coeff, group=Sample), stat="density", alpha=0.1) + 
  geom_density(data=dat.mu, aes(x=log.est), color='red') + 
  xlim(c(-0.06, .06))

```

@@
